---
title: "NHP ratio reference values calculation"
author: "Rodrigo Arcoverde Cerveira & Xianglei Yan"
date: '`r format(Sys.Date(), "%Y-%m-%d")`'
output: html_document
knit: (function(inputFile, encoding) {
          rmarkdown::render(inputFile,
                            encoding = encoding, 
                            output_file = paste0(
                              xfun::sans_ext(inputFile), '_', Sys.Date(), '.html'),
                                output_dir = "../results/lab_book/")})
editor_options: 
  chunk_output_type: inline
---


```{r, global.options, include=FALSE}
# Global knitr settings 
# Set up figures and chunks messages

knitr::opts_knit$set(
  echo = TRUE,
  root.dir = getwd(),
  fig.width = 6, fig.height = 5,
  fig.align = "center",
  out.width = 768,
  fig.pos = "H",
  warning = FALSE,
  message = FALSE
)
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.width = 6, fig.height = 5,
  fig.align = "center",
  out.width = 768,
  fig.pos = "H"
)

data.dir <- "../data"
result.dir <- paste("results/", Sys.Date(), sep = "/") 
figure.dir <- paste0("../", result.dir, "/figures") 

# Create result.dir with date in if not existent
ifelse(isFALSE(dir.exists(paste0("../", result.dir))),
  dir.create(paste0("../", result.dir), recursive = TRUE),
  "Result directory for today exists already!"
)

ifelse(isFALSE(dir.exists(paste0(figure.dir))),
  dir.create(paste0(figure.dir), recursive = TRUE),
  "Result directory for today exists already!"
)

options(stringsAsFactors = FALSE)


```

# Load libraries
```{r, include=FALSE, results='hide'}
# Load libraries
library(data.table)
library(dplyr)
library(readxl)
library(tidyr)
library(ggplot2)
library(ggprism)
```

# Load data

```{r load_data}
# load all excel files within data folder with "xlsx" file extension
ls_excel <- list.files(data.dir, recursive = TRUE, all.files = TRUE, full.names = TRUE, pattern = "xlsx")[-11]
names(ls_excel) <- list.files(data.dir[-11], recursive = TRUE, all.files = TRUE, pattern = "xlsx")[-11]
ls_excel <- lapply(ls_excel, read_xlsx)

# merge all files in a single data.table and format column names
ls_excel <- rbindlist(ls_excel) %>% 
  select(-c(...13,...14)) %>%
  rename("standard_10^9/L" = "10^9/L")
```

# Editing to a wider data.table

```{r data_wrangling}
clean_dataset <- ls_excel %>%
  pivot_wider(names_from = c("measurement_name"), 
              values_from = c("standard_10^9/L", "measurement_value")) %>%
  drop_na(any_of(names(.)[contains("measurement_value")]))

```

## Calculating ratios

This section calculates four important blood cell ratios used in hematology:
 1. NLR (Neutrophil-to-Lymphocyte Ratio) - inflammation marker
 2. MLR (Monocyte-to-Lymphocyte Ratio) - immune system indicator
 3. PLR (Platelet-to-Lymphocyte Ratio) - inflammation and thrombosis marker
 4. NMR (Neutrophil-to-Monocyte Ratio) - immune response indicator

The calculated ratios are added as new columns to the dataset.

```{r ratio_calc}

clean_dataset <- clean_dataset %>%
  mutate(NLR = (`measurement_value_Neutrophil-Band-Abs`+`measurement_value_Neutrophil-Seg-Abs`)/ `measurement_value_Lymphocytes-Abs`,
         MLR = `measurement_value_Monocytes-Abs`/`measurement_value_Lymphocytes-Abs`,
         PLR = measurement_value_Platelets/`measurement_value_Lymphocytes-Abs`,
         NMR = (`measurement_value_Neutrophil-Band-Abs`+`measurement_value_Neutrophil-Seg-Abs`)/`measurement_value_Monocytes-Abs`
         )

write.csv(clean_dataset, file = paste0("../", result.dir, "/merged_dataset_ratios.csv"), 
          row.names = F)
```

## Simple visualization

This section creates violin plots to visualize the distribution of blood cell ratios:
 1. First transforms the data from wide to long format for plotting
 2. Creates age groups (Juvenile, Young Adults, Adults, Middle-Aged, Elderly)
 3. Generates separate plots for each age group using facet_wrap
 4. Uses log10 scale for better visualization of ratio distributions
 5. Colors plots by species for comparison
 6. Saves the final plot as a TIFF image in the figures directory

```{r plot}

clean_dataset %>%
  pivot_longer(cols = c(NLR, MLR, PLR, NMR), values_to = "ratio_value", names_to = "ratio") %>%
  mutate(age_group = factor(case_when(age <= 3 ~ "Juvenile",
                               age > 3 & age <= 7 ~ "Young Adults",
                               age > 7 & age <= 12 ~ "Adults",
                               age > 12 & age <= 17 ~ "Middle-Aged",
                               age > 17 ~ "Elderly"
                               ), levels = c("Juvenile", "Young Adults", "Adults", "Middle-Aged", "Elderly") )) %>%
  ggplot(aes(x = ratio, y = ratio_value, color = species_common_name)) +
  geom_violin(trim = FALSE, scale = "area", draw_quantiles = TRUE) +
  scale_y_log10() + 
  labs( y = "Ratio values (log10)", x = "", color = "") +
  facet_wrap(~age_group) +
  theme_classic() +
  theme(aspect.ratio = 1)

ggsave(paste0(figure.dir,"/ratios_plot_per_spescies.tiff"), height = 5)
```

```{r}
sessionInfo()
```